Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-19 paper/paper.tex
edits in the introduction section. Tested everything on bennet.

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of
division~\cite{adams2009bacterial,lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD
oscillation.~\cite{hu1999topological,fu2001mine,shapiro2009and,yu1999ftsz,raskin1999rapid,meacci2005min,raskin1999minde}.

Previous studies have shown that the minD protein system is capable of
exhibiting oscillations in round shapes \cite{fange2006noise}as well
as in connected three pronged tube shapes~\cite{varma2008min}, in
which the oscillations seem to seek out the extreme poles in the
cell~\cite{corbin2002exploring,juarez2010changes}.

However, Mannik \emph{et al.} have also shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2010bacteria,mannik2009bacterial}. They have
experimentally forced E. Coli cells into silicon microfabricated
channels of $0.25\micron$ thickness. Upon entering the channels, the
cells undergo a rapid mechanical deformation in which they widen along
their short axis parallel to the channel plane, followed by a slower,
broadening out along both free axes.  This deformation results in very
wide (they can reach widths of over $5\micron$), flattened cells that
when view from the top down show very irregular and assymetric shapes.
While these cells are still able to divide into surprisingly equal
volumes, the MinD oscillations within lose their regularity, both
temporally and spatially. Seen with flourescent microscopy (after
using common genetic techniques to allow the cells to flouresce), the
MinD seem to maximize in multiple locations within the cell, in a
seemingly random sequence. This allows for an opportunity to test
previously studied MinD simulation models against more extreme
experimental cases than have been seen thus far. \fixme{Perhaps this
  sentence should come later after we've introduced the simulation
  research that's been done.}

A significant amount of work has been done to develop protein reaction
and diffusion simulation models that exhibit accurate macroscopic
dynamics of the MinD protein system. Early models involve free
proteins that affect each others' rates of diffusion and membrane
attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is more
likely to attach to the membrane when MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\cite{hsieh2010direct,mileykovskaya2003effects}.

Variations of the Huang 2003 model that stochasticly account for
molecular interaction and diffusion~\cite{fange2006noise} and
monte-carlo simulations that implement this stochastic version of
Huang's mean field reaction rates~\cite{kerr2006division} broadly
confirm the major results obtained by Huang's model.  However, they
more successfully predict experimentally observed oscillations in
round cell phenotypes~\cite{fange2006noise, huang2004min} and
variations in other
quantities~\cite{kruse2007experimentalist}\fixme{If going to use this
  sentence should list these and reference properly?}. Both
deterministic and stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial,robb2014stochastic,oguz2014stochastic,fu2013deterministic,rudiger2014stochastic},
and it's important to better understand their limitations.  Manniks'
experiments allow for a new distinct opportunity to test these two
approaches further.
%% Biochemical models of broader scope have also been
%% used to study the MinD system and show consistent
%% results~\cite{arjunan2010new}.

@@

In this paper we apply a deterministic model and a stochastic model
of simulation and compare them against Mannik's experimental findings.

\fixme{Following paragraph is awkward, and I don't think it's needed.}
Studies have shown as well that MinD binds preferentially to regions
enriched with cadiolipin, an anionic phospholipid that collects on
regions of high negative curvature. This mechanism has been
incorporated into other
models.\cite{drew2005polymerization,cytrynbaum2007multistranded,renner2012mind,renner2012mind}
However, this mechanism of combined clustering, phospholipids and MinD
has not been observed in real cells. \cite{halatek2012highly}

>>>>

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD oscillation.~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid,
  meacci2005min, raskin1999minde}.

Previous studies have shown experimentally that the MinD protein
system is capable of exhibiting oscillations in round
shapes~\cite{fange2006noise} as well as in connected three pronged
tube shapes~\cite{varma2008min}, in which the oscillations seem to
seek out the extreme poles in the cell~\cite{corbin2002exploring,
  juarez2010changes}.  One conclusion of these studies has been that
the MinD system robustly forms bipolar oscillations regardless of
variations in cellular shape.

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2010bacteria, mannik2009bacterial}. They have
experimentally forced E. Coli cells into silicon microfabricated
channels of $0.25\micron$ thickness. Upon entering the channels, the
cells undergo a mechanical deformation in which they both widen and
lengthen within the plane of the channel.  This deformation results in
very wide (they can reach widths of over $5\micron$) flattened cells
that when viewed from the top down show irregular and asymmetric
shapes.  While these cells are still able to divide into surprisingly
equal volumes, the MinD oscillations in these cells is irregular,
both temporally and spatially. Seen with flourescent microscopy (after
using common genetic techniques to allow the cells to flouresce), the
MinD seem to maximize in multiple locations within the cell, in a
seemingly random sequence. This allows for an opportunity to test
previously studied MinD simulation models against more extreme
experimental cases than have been seen thus far.

A number of models of the MinD protein system have been developed,
which accurately describe the dynamics using simulations of protein
reaction and diffusion.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
work with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\fixme{finding what results? We don't want to just list what has
  been done, but rather to summarize the state of
  knowledge.}~\cite{hsieh2010direct, mileykovskaya2003effects}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically treat molecular interaction
and diffusion~\cite{fange2006noise, kerr2006division}.  These studies
largely confirm the results of Huang's model, when applied to the
wild-type phenotype.  However, the stochastic models more successfully
predict experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and enable prediction
of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

In this paper we use Huang's 2003 model~\cite{huang2003dynamic} in
both deterministic and stochastic variants to study flattened cells
similar to those observed by Mannik~\cite{mannik2009bacterial}.
